Normalization by Evaluation for Sized Dependent Types

### 2017-06-10 nbe/icfp17.tex
[ ICFP 17 ] No longer erase to infty during readback.

\noindent
\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}\ \ Read back unreflected neutral.
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  \ru{
    }{\RElim k {\ann \alpha} {\ann \infty}}
  % \ru{\RSize k \alpha b
  %   }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\R k D V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] D {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\R k D V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] D d)} {\fix[\ell] V v}}
\end{gather*}

@@

The PERs $\NE$ characterizes neutral reifiable terms.
% \[
% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
%     & \defas &
%     T \evalsto N \mand T' \evalsto N' \mand
%     \mforall k \mboth
%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \nfeq m'
%     .\\
% \end{array}
% \]
The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
\[
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
  n = n' \in \NE
    & \defas &
    \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
  d = d' \in \NF
    & \defas &
    \R k d v &\mand& \R k {d'} {v} &\mforall k \\
  e = e' \in \ELIM
    & \defas &
    \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
  A = A' \in \TY
    & \defas &
    \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
  % \fbox{$n = n' \in \NE$}
  %   & \defas &
  %   \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
  % \fbox{$d = d' \in \NF$}
  %   & \defas &
  %   \R k d v &\mand& \R k {d'} {v} &\mforall k \\
  % \fbox{$e = e' \in \ELIM$}
  %   & \defas &
  %   \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
  % \fbox{$A = A' \in \TY$}
  %   & \defas &
  %   \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
\end{array}
% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$n = n' \in \NE$}
%     & \defas &
%     \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m \nfeq m' &\mforall k \\
%   \fbox{$d = d' \in \NF$}
%     & \defas &
%     \R k d v &\mand& \R k {d'} {v'} &\mand& v \nfeq v' &\mforall k \\
%   \fbox{$e = e' \in \ELIM$}
%     & \defas &
%     \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v \nfeq e_v' &\mforall k \\
%   \fbox{$A = A' \in \TY$}
%     & \defas &
%     \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V \nfeq V' &\mforall k \\
% \end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \nfeq n'
  %   }{t = t' \in \NE}
\]
%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

@@

\item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
$v_1,v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
It is sufficient to show
$\R{k+1}{\down{B_i[u_i]}{(t_i\,u_i)}} {v_i}$.
By definition of $\EL(T)$ we have $t_1\,u_1 = t_2\,u_2 \in B[u_1]$, thus, by induction hypothesis,
$\down{B_1[u_1]}(t_1\,u_1) = \down{B_2[u_2]}(t_2\,u_2) \in \NF$, which delivers $v_1$ and $v_2$ for $k+1$.
\end{enumerate}

\nextcase $T \evalsto \forallT B$ with $B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE$
\begin{gather*}
%   \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
%     }{T \in \SET[\ell]}
% \\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_1}
%\\[2ex]
\qquad
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.
\qed
\end{enumerate}

@@

\begin{theorem}[Completeness of NbE]
  If\/ $\Gamma \der t = t' : T$ then there is a normal form $v$ such that
  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v$.
  % If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
  % $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem, $T \rho_\Gamma \in \SET[\ell]$ for some $\ell$ and
  $t \rhog = t' \rhog \in T\rhog$.  By reification (Cor.~\ref{cor:rere}) we have
  $\down{T\rhog} (t\rhog) = \down{T\rhog}(t'\rhog) \in \NF$ which implies the theorem
  by read back with $k = |\Gamma|$.
\end{proof}

@@

\begin{corollary}[Soundness of NbE]
  \label{cor:soundnbe} \bla
  \begin{enumerate}
  \item \label{it:sound}
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
  \item
  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \equiv \nbe \Gamma T {t'}$ then
  $\Gamma \der t = t' : T$.
  \end{enumerate}
\end{corollary}
\begin{proof*}
\bla
\begin{enumerate}

@@

\item By transitivity of judgmental equality from (\ref{it:sound}).
   Here we use that the computed normal forms are identical because
   we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
\qed
\end{enumerate}
\end{proof*}

\begin{corollary}[Decidability of judgemental equality]
  If\/ $\Gamma \der t,t' : T$ then the test whether $\nbe\Gamma T t \equiv \nbe\Gamma T{t'}$
  terminates and
  decides $\Gamma \der t = t' : T$.
\end{corollary}
\LONGVERSION{
\begin{proof}
  Follows directly from soundness (which includes termination) and completeness of NbE.
\end{proof}
} % END LONGVERSION

@@

Reification erases all irrelevant size arguments to $\infty$.  We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.  However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.  Proving this statement seems to be the key to an incremental equality check; but
the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.

>>>>

\noindent
\fbox{$\RNe k n m$} and \fbox{$\RElim k e e^v$}\ \ Read back unreflected neutral.
\begin{gather*}
%   \ru{t \evalsto \lev j
%     }{\RNe k t {\ind{k-1-j}}}
% \qquad
  \ru{%t \evalsto \lev j\,\vec e \qquad
      \RElim k {e_i} {e^v_i} \mforall i
    }{\RNe k {\lev j\,\vec e} {\ind{k \monus (1+j)}\,\vec e^v}}
\qquad
% \end{gather*}
% \begin{gather*}
  \ru{\R k d v
    }{\RElim k d v}
\qquad
  \ru{\RSize k \alpha b
    }{\RElim k \alpha b}
\qquad
  % \ru{
  %   }{\RElim k {\ann \alpha} {\ann \infty}}
  \ru{\RSize k \alpha b
    }{\RElim k {\ann \alpha} {\ann b}}
\\[2ex]
  \ru{\R k D V \qquad
      \R k {d_z} {v_z} \qquad
      \R k {d_s} {v_s}
    }{\RElim k {(\case[\ell] D {d_z} {d_s})} {\case[\ell] V {v_z} {v_s}}}
\qquad
  \ru{\R k D V \qquad
      \R k d v
    }{\RElim k {(\fix[\ell] D d)} {\fix[\ell] V v}}
\end{gather*}

@@

%The PERs $\NE$ characterizes neutral reifiable terms.
% \[
% \begin{array}{lll}%{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   \fbox{$\up T n = \up {T'} {n'} \in \NE$}
%     & \defas &
%     T \evalsto N \mand T' \evalsto N' \mand
%     \mforall k \mboth
%     \RNe k n m \mand \RNe k {n'} {m'} \mand m \nfeq m'
%     .\\
% \end{array}
% \]
The PERs $\NE$ and $\NF$ characterize (neutral) normalizing terms.
\[
% \begin{array}{l@{\quad}l@{\quad}l@{~}l@{~}l@{~}l@{~}l@{~}l}
%   n = n' \in \NE
%     & \defas &
%     \RNe k n m &\mand& \RNe k {n'} {m} &\mforall k \\
%   d = d' \in \NF
%     & \defas &
%     \R k d v &\mand& \R k {d'} {v} &\mforall k \\
%   e = e' \in \ELIM
%     & \defas &
%     \RElim k e e_v &\mand& \R k {e'} {e_v} &\mforall k \\
%   A = A' \in \TY
%     & \defas &
%     \RTy k A V &\mand& \RTy k {A'} {V} &\mforall k \\
% \end{array}
%\newcommand{\myfbox}[1]{\begin{array}[c]{@{}l@{}}\framebox(60,12)[l]{~\,#1}\end{array}}
\newcommand{\myfbox}[1]{\framebox(60,12)[l]{~\,#1}}
\begin{array}{l@{\quad}l@{\quad}l@{~}l@{\quad}l@{~}l@{~\;}r@{~}l@{~}l@{~}l}
  \raisebox{-0.5ex}{\myfbox{$n = n' \in \NE$}}
    & \defas &
    \RNe k n m &\mand& \RNe k {n'} {m'} &\mand& m &\nfeq& m' &\mforall k
\\[1ex]
  \raisebox{-0.8ex}{\myfbox{$d = d' \in \NF$}}
    & \defas &
    \R k d v &\mand& \R k {d'} {v'} &\mand& v &\nfeq& v' &\mforall k
\\[1ex]
  \raisebox{-0.6ex}{\myfbox{$e = e' \in \ELIM$}}
    & \defas &
    \RElim k e e_v &\mand& \R k {e'} {e_v'} &\mand& e_v &\nfeq& e_v' &\mforall k
\\[1ex]
  \raisebox{-0.8ex}{\myfbox{$A = A' \in \TY$}}
    & \defas &
    \RTy k A V &\mand& \RTy k {A'} {V'} &\mand& V &\nfeq& V' &\mforall k \\
\end{array}
  % \ru{\RNe k t m \mand \RNe k {t'} {m'} \mand m \nfeq m' \mforall k
  %   }{t = t' \in \NE}
  % \ru{t \evalsto \up N n \qquad t' \evalsto \up{N'}{n'} \qquad n \nfeq n'
  %   }{t = t' \in \NE}
\]
%All PERs $\A$ we consider are sandwiched $\NE \subseteq \A \subseteq \NF$ between $\NE$ and $\NF$.

@@

\item We assume $k \in \NN$ and show $\R k {\down{S_i}{t_i}} \lambda v_i$ for some normal forms
$v_1 \nfeq v_2$.  Let $u_i = \up{A_i} \lev k$.  Note that $u_1 = u_2 \in A$ by induction hypothesis.
It is sufficient to show
$\R{k+1}{\down{B_i[u_i]}{(t_i\,u_i)}} {v_i}$.
By definition of $\EL(T)$ we have $t_1\,u_1 = t_2\,u_2 \in B[u_1]$, thus, by induction hypothesis,
$\down{B_1[u_1]}(t_1\,u_1) = \down{B_2[u_2]}(t_2\,u_2) \in \NF$, which delivers $v_1$ and $v_2$ for $k+1$.
\end{enumerate}

\nextcase $T \evalsto \forallT B$ with $B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE$
\begin{gather*}
%   \ru{B[\alpha] \in \SET[\ell] \mforall \alpha \in \SIZE
%     }{T \in \SET[\ell]}
% \\[2ex]
  \ru{S_1 \evalsto \forallT{B_1} \qquad
      B[\alpha] \shape B_1[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_1}
%\\[2ex]
\qquad
  \ru{S_2 \evalsto \forallT{B_2} \qquad
      B[\alpha] \shape B_2[\alpha'] \mforall \alpha, \alpha' \in \SIZE
    }{T \shape S_2}
\end{gather*}
\begin{enumerate}
\item To show $\up{S_1}{n_1} = \up{S_2}{n_2} \in T$ assume arbitrary $\alpha_1,\alpha_2 \in \SIZE$.
  Since $n_1 \ann{\alpha_1} = n_2 \ann{\alpha_2} \in \NE$ by Lemma~\ref{lem:closne}, we obtain
  $\up{B_1[\alpha_1]}{(n_1 \ann{\alpha_1})} = \up{B_2[\alpha_2]}{(n_2 \ann{\alpha_2})} \in B[\alpha_1]$
  by induction hypothesis.
  Thus, $(\up{S_1}{n_1}) \ann{\alpha_1} = (\up{S_2}{n_2}) \ann{\alpha_2} \in B[\alpha_1]$
  by weak head expansion, which entails the goal by definition of $\EL(T)$.
\item Assume $k \in \NN$ and note that $\lev k = \lev k \in \SIZE$, hence,
  $t_1\,\ann{\lev k} = t_2\,\ann{\lev k} \in B[\lev k]$.  Thus, by induction hypothesis,
  $\R{k+1}{\down{B_i[\lev k]}(t_i\,\ann{\lev k})}{v_i}$ with $v_1 \nfeq v_2$, and finally
  $\R k {\down{S_i} t_i} {\lambda v_i}$ by definition of read back.
\qed
\end{enumerate}

@@

\begin{theorem}[Completeness of NbE]
  % If\/ $\Gamma \der t = t' : T$ then there is a normal form $v$ such that
  % $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v$.
  If\/ $\Gamma \der t = t' : T$ then there are normal forms $v \nfeq v'$ such that
  $\nbe \Gamma T t \evalsto v$ and $\nbe \Gamma T {t'} \evalsto v'$.
\end{theorem}
\begin{proof}
  By the fundamental theorem, $T \rho_\Gamma \in \SET[\ell]$ for some $\ell$ and
  $t \rhog = t' \rhog \in T\rhog$.  By reification (Cor.~\ref{cor:rere}) we have
  $\down{T\rhog} (t\rhog) = \down{T\rhog}(t'\rhog) \in \NF$ which implies the theorem
  by read back with $k = |\Gamma|$.
\end{proof}

@@

\begin{corollary}[Soundness of NbE]
  \label{cor:soundnbe} \bla
  \begin{enumerate}
  \item \label{it:sound}
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
  \item
  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \nfeq \nbe \Gamma T {t'}$ then
  $\Gamma \der t = t' : T$.
  \end{enumerate}
\end{corollary}
\begin{proof*}
\bla
\begin{enumerate}

@@

\item By transitivity of judgmental equality from (\ref{it:sound}), using
  Lemma~\ref{lem:erase}:
  $\Gamma \der t = \nbe \Gamma T t = \nbe \Gamma T {t'} = t' : T$.
   % Here we use that the computed normal forms are identical because
   % we erase all size annotations $\ann\alpha$ to $\ann\infty$ during read back.
\qed
\end{enumerate}
\end{proof*}

\begin{corollary}[Decidability of judgemental equality]
  If\/ $\Gamma \der t,t' : T$ then the test whether $\nbe\Gamma T t \nfeq \nbe\Gamma T{t'}$
  terminates and
  decides $\Gamma \der t = t' : T$.
\end{corollary}
\LONGVERSION{
\begin{proof}
  Follows directly from soundness (which includes termination) and completeness of NbE.
\end{proof}
} % END LONGVERSION

@@

% Reification erases all irrelevant size arguments to $\infty$.  We have also considered to keep the sizes there, and define equality $v \approx v'$ of normal forms up to differences in erased arguments.  However, it seems difficult to prove that equal normal forms are judgmentally equal, \ie, that $\Gamma \der v,v' : T$ and $v \approx v'$ imply $\Gamma \der v = v' : T$.  Proving this statement seems to be the key to an incremental equality check; but
% the usual bidirectional argument \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.
In the long run, we wish for an incremental type-directed equality check, which is also at the heart of Agda's type checker, as this check can generate constraints for meta variables involved in type reconstruction \cite{norell:PhD}.
However, the usual bidirectional construction \cite{abelScherer:types10} does not seem to go through as we lack uniqueness of types.

### 2017-06-10 nbe/sec-asub.tex
[ ICFP 17 ] No longer erase to infty during readback.

Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
\begin{gather*}
  \ru{T \evalsto n \qquad
      T' \evalsto n' \qquad
      \Nbe \Gamma n \equiv \Nbe \Gamma n'
    }{\Gamma \der T \asub T'}
\\[2ex]
  \rux{T \evalsto \Set[\ell] \qquad
      T' \evalsto \Set[\ell']
    }{\Gamma \der T \asub T'
    }{\ell \leq \ell'}
\qquad
  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
      T_2' \evalsto \piT{U_2}{T_2} \qquad
      \Gamma \der U_2 \asub U_1 \qquad
      \cext \Gamma {U_2} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\\[2ex]
  \ru{T \evalsto \Nat a \qquad
      T' \evalsto \Nat a'
    }{\Gamma \der T \asub T'
    }{a \leq a'}
\qquad
  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
      T_2' \evalsto \epiT\Size{T_2} \qquad
      \cext \Gamma {\Size} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\end{gather*}

@@

\nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \equiv \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.

>>>>

Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
\begin{gather*}
  \ru{T \evalsto n \qquad
      T' \evalsto n' \qquad
      \Nbe \Gamma n \nfeq \Nbe \Gamma n'
    }{\Gamma \der T \asub T'}
\\[2ex]
  \rux{T \evalsto \Set[\ell] \qquad
      T' \evalsto \Set[\ell']
    }{\Gamma \der T \asub T'
    }{\ell \leq \ell'}
\qquad
  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
      T_2' \evalsto \piT{U_2}{T_2} \qquad
      \Gamma \der U_2 \asub U_1 \qquad
      \cext \Gamma {U_2} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\\[2ex]
  \ru{T \evalsto \Nat a \qquad
      T' \evalsto \Nat a'
    }{\Gamma \der T \asub T'
    }{a \leq a'}
\qquad
  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
      T_2' \evalsto \epiT\Size{T_2} \qquad
      \cext \Gamma {\Size} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\end{gather*}

@@

\nextcase $A_i' \evalsto N_i$ and $T_i' \evalsto n_i$ and $N_1 = N_2 \in \NE$ and $\RTY \Gamma {n_i} {N_i} {s_i}$.
By Cor.~\ref{cor:oneone} $\Gamma \der n_1 = n_2$, thus, by completeness of NbE, $\Nbe\Gamma n_1 \nfeq \Nbe\Gamma n_2$, which entails $\Gamma \der T_1' \asub T_2'$.

### 2017-06-10 nbe/sec-sr.tex
[ ICFP 17 ] No longer erase to infty during readback.

For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
Analogously, we show that the test $\Nbe \Gamma T \equiv \Nbe \Gamma T'$
decides type equality $\Gamma \der T = T'$.

@@

Last, we prove statement (\ref{it:piinj}), the last statement follows analogously.
By the fundamental theorem,
$(\piT U T)\rhog = (\piT{U'}{T'})\rhog \in s$ which by definition means
$U\rhog = U'\rhog \in s$ and $T(\rhog,u) = T'(\rhog,u') \in s$ for all $u = u' \in s$.
By reification (Thm.~\ref{thm:rere}),
$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V$ for some normal form
$V$.
Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
Now, choosing $u := u' := \up {U\rhog} \var\Gamma \in U$, we obtain
$T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
\end{proof}
} % END LONGVERSION
\begin{lemma}[Type constructor discrimination]
  \label{lem:discr}
  Different type constructors are not related by subtyping.  For instance,
  $\Gamma \der \Nat a \leq \piT U T$ and
  $\Gamma \der \erpiT \Size T \leq \piT \Size T'$ are both impossible.
%  There are $4\times4 = 16$ statements of that sort.
  % The following judgments are \emph{not} derivable.
  % \begin{enumerate}
  % \item $\Gamma \der \Nat a \leq \Set[\ell]$ and $\Gamma \der \
  % \end{enumerate}
\end{lemma}
\begin{proof}
  Follows directly by the fundamental theorem (Thm.~\ref{thm:fund}) and inversion on semantic subtyping.
\end{proof}
% \section{Algorithmic Subtyping}
% \label{sec:algosub}

>>>>

For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
Analogously, we show that the test $\Nbe \Gamma T \nfeq \Nbe \Gamma T'$
decides type equality $\Gamma \der T = T'$.

@@

Last, we prove statement (\ref{it:piinj}), the last statement follows analogously.
By the fundamental theorem,
$(\piT U T)\rhog = (\piT{U'}{T'})\rhog \in s$ which by definition means
$U\rhog = U'\rhog \in s$ and $T(\rhog,u) = T'(\rhog,u') \in s$ for all $u = u' \in s$.
By reification (Thm.~\ref{thm:rere}),
$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhog} V'$ for some normal forms
$V \nfeq V'$.
Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
Now, choosing $u := u' := \up {U\rhog} \var\Gamma \in U$, we obtain
$T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
\end{proof}
} % END LONGVERSION
\begin{lemma}[Type constructor discrimination]
  \label{lem:discr}
  Different type constructors are not related by subtyping.  For instance,
  $\Gamma \der \Nat a \leq \piT U T$ and
  $\Gamma \der \erpiT \Size T \leq \piT \Size T'$ are both impossible.
%  There are $4\times4 = 16$ statements of that sort.
  % The following judgments are \emph{not} derivable.
  % \begin{enumerate}
  % \item $\Gamma \der \Nat a \leq \Set[\ell]$ and $\Gamma \der \
  % \end{enumerate}
\end{lemma}
\begin{proof}
  Follows directly by the fundamental theorem (Thm.~\ref{thm:fund}) and inversion on semantic subtyping.
\end{proof}
% \section{Algorithmic Subtyping}
% \label{sec:algosub}
